Exploring the Traveling Salesman Problem

### 2017-05-03 paper/paper.tex
update

% Packages
% ---
\usepackage[
  left=3cm,
  right=3cm,
  top=2cm,
  bottom=2cm]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{clrscode3e}
\usepackage{blkarray}
\usepackage{booktabs}
\usepackage{parskip}
\usepackage{mathptmx}
\usepackage{bigstrut}
\usepackage{mathptmx}
\usepackage{minted}
\usepackage[pdftex,pdfpagelabels,bookmarks,hyperindex,hyperfigures]{hyperref}
\usepackage{apacite}
\usepackage{tikz}

@@

\subsection{Held--Karp}
The Held--Karp algorithm is an $O(2^n)$ approach to solving the TSP\@.
It is not fast enough to solve TSP instances with a large number of
cities, but the solution for the instances it can solve is guarenteed
to be correct.  The algorithm works as follows:
\par

@@

The last step holds since $G$ satisfies the triangle inequality.

@@

@@

Christofide's algorithm and the Nearest Neighbor algorithm both perform at an
acceptable level (that is, they return the optimal tour enough of the time) for
Euclidean TSP instances.  If a TSP instance is Euclidean, then it satisfies the
triangle inequality.  This property is necessary for Christofide's algorithm to
return the correct tour \cite{nicos} and it greatly increases the probability
that the Nearest Neighbor algorithm will as well \cite{gutin}.
\par

For asymmetric and general symmetric TSP instances, Christofide's algorithm will
never return the optimal tour.  The Nearest Neighbor algorithm, on the other hand,
can find an optimal tour, but it's highly unlikely.  Additionally, for every $n\gte 2$
there exists at least one asymmetric or general symmetric TSP instance for which the
Nearest Neighbor algorithm returns the \textit{worst} tour. \cite{gutin}.  We are
optimizing for correctness over speed and problem size in this paper and will continue
to explore the Held--Karp algorithm.

@@

\subsection{Held-Karp Runtime}
We will show that the Held-Karp algorithm is an improvement over the brute force
approach by proving the runtime of Held-Karp is $O(n^{2}2^n)$, which is better than
the runtime of brute force, $O(n!)$.  Observe from the pseudocode that our algorithm
involves computing every possible subset of $\{1,\dots,n\}$, where $n$ is the number
of cities.  Knowing this, we can prove the runtime by proving a set with $n$ elements
has $2^n$ subsets using induction.

@@

% Appendicies
% ---
%\newpage
%\appendix
%\section{Appendices}
%\addcontentsline{toc}{section}{Appendices}
%\renewcommand{\thesubsection}{\Alph{subsection}}
%
%\subsection{Collaboration}
%  I collaborated with Grant Eaton and Luke Artnak.
%
%\newpage
%\subsection{Code Demonstration}
%  \inputminted[frame=lines,linenos=true]{python}{../src/brute_force.py}
%
\newpage
\bibliographystyle{apacite}
\bibliography{biblio}

>>>>

% Packages
% ---
\usepackage[
  left=3cm,
  right=3cm,
  top=2cm,
  bottom=2cm]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{clrscode3e}
\usepackage{blkarray}
\usepackage{booktabs}
\usepackage{parskip}
\usepackage{mathptmx}
\usepackage{bigstrut}
\usepackage{caption}
\usepackage{mathptmx}
\usepackage{minted}
\usepackage[pdftex,pdfpagelabels,bookmarks,hyperindex,hyperfigures]{hyperref}
\usepackage{apacite}
\usepackage{tikz}
\usepackage{float}
\usepackage{graphicx}

@@

\subsection{Held--Karp}
The Held--Karp algorithm is an $O(n^2 2^n)$ approach to solving the TSP\@.
It is not fast enough to solve TSP instances with a large number of
cities, but the solution for the instances it can solve is guarenteed
to be correct.  The algorithm works as follows:
\par

@@

The last step holds since $G$ satisfies the triangle inequality. \cite{nicos}

@@

This is an $O(n^2)$ algorithm because on each iteration the
current city visits every other unvisited city.

@@

Christofide's algorithm and the Nearest Neighbor algorithm both perform at an
acceptable level (that is, they return the optimal tour enough of the time) for
Euclidean TSP instances \cite{gutin}.  If a TSP instance is Euclidean, then it satisfies the
triangle inequality.  This property is necessary for Christofide's algorithm to
return the correct tour \cite{nicos} and it greatly increases the probability
that the Nearest Neighbor algorithm will as well \cite{gutin}.
\par

For asymmetric and general symmetric TSP instances, Christofide's algorithm will
never return the optimal tour.  The Nearest Neighbor algorithm, on the other hand,
can find an optimal tour, but it's highly unlikely.  Additionally, for every $n \geq 2$
there exists at least one asymmetric or general symmetric TSP instance for which the
Nearest Neighbor algorithm returns the \textit{worst} tour. \cite{gutin}.  We are
optimizing for correctness over speed and problem size in this paper and will continue
to explore the Held--Karp algorithm.

@@

\subsection{Held-Karp Runtime}
We will show that the Held-Karp algorithm is an improvement over the brute force
approach by proving the runtime of Held-Karp is $O(n^{2}2^n)$, which is better than
the runtime of brute force, $O(n!)$.

Observe from the pseudocode that our algorithm iterates from $2$ to $n$.  In that
loop we also iterate from $1$ to $n$.  This is where the $n^2$ portion of our runtime
comes from.  Also observe that the algorithm involves computing every possible subset
of $\{1,\dots,n\}$, where $n$ is the number of cities.  Knowing this, we can prove the
runtime by proving a set with $n$ elements has $2^n$ subsets using induction.

@@

\section{Experimental Comparison}

\subsection{Overview}
We will do an experiment on the brute force and Held--Karp algorithms,
running Python implementations of both algorithms and comparing their
execution times.

\subsection{Input}
The input for both algorithms will be a standard TSPLIB sample instance, as shown
in the table below.  The distance matrix is created using the $x$ and $y$ coordinates
in the table and is used by both algorithms to generate the optimal tour.
\par

\begin{table}[H]
  \centering
  \caption*{TSPLIB xqf131 input}
  \begin{tabular}{c|c|c}
    \toprule
    City & $x$ & $y$ \\
    \midrule
    1& 0& 13 \\
    2& 0& 26 \\
    3& 0& 27 \\
    4& 0& 39 \\
    5& 2& 0\\
    6& 5& 13\\
    7& 5& 19\\
    8& 5& 25\\
    9& 5& 31\\
    10& 5& 37\\
    11& 5& 43\\ 
    12& 5& 8\\ 
    13& 8& 0\\
    14& 9& 10\\
    15& 10& 10\\
    16& 11& 10\\
    17& 12& 10\\
    18& 12& 5\\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Execution}
We run both algorithms on the data set above and produce the runtime table
below.  Note that the brute force times for $n>10$ are estimated as the
algorithm takes too long to run for those problem sizes.
\par

\begin{table}[H]
  \centering
  \caption*{Runtime comparison}
  \begin{tabular}{c|c|c|c}
    \toprule
    Cities $n$ & Brute Force (s) & Held-Karp (s) & BF/HC \\
    \midrule
    4   &       0.00003       &    0.00005    &      0.60748      \\
    5   &       0.00010       &    0.00011    &      0.89353      \\
    6   &       0.00062       &    0.00032    &      1.94113      \\
    7   &       0.00437       &    0.00091    &      4.77795      \\
    8   &       0.03484       &    0.00226    &      15.37989     \\
    9   &       0.29779       &    0.00568    &      52.38270     \\ 
    10  &       2.84006       &    0.01420    &     199.93426     \\
    11  &       31.24068      &    0.03465    &     901.55505     \\
    12  &      374.88817      &    0.09049    &    4,143.09304    \\
    13  &     4,873.54623     &    0.24270    &    20,080.92283   \\
    14  &     68,229.64716    &    0.60574    &   112,639.24419   \\
    15  &   1,023,444.70737   &    1.44135    &   710,060.87550   \\
    16  &   16,375,115.31784  &    3.43850    &  4,762,288.58166  \\
    17  &  278,376,960.40329  &    8.09218    &  34,400,754.72416 \\
    18  & 5,010,785,287.25922 &    19.04262   & 263,135,354.10685 \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Output}
The output from running the Held-Karp algorithm for 18 cities is a
list representing the optimal path:
$(1, 2, 3, 4, 11, 10, 9, 8, 7, 6, 14, 15, 16, 17, 18, 13, 5, 12, 1)$

Our program can also produce a graph:

\centering
\includegraphics[scale=0.5]{plot.png}

\newpage
\appendix
\section*{Appendices}
\addcontentsline{toc}{section}{Appendices}
\renewcommand{\thesubsection}{\Alph{subsection}}

\subsection{Collaboration}
  I collaborated with Grant Eaton and Luke Artnak.

\newpage
\subsection{Code Demonstration}
  \inputminted[frame=lines,linenos=true]{python}{../src/brute_force.py}

\newpage
\bibliographystyle{apacite}
\bibliography{biblio}
